<!--
%\VignetteIndexEntry{01.1 Introduction to R}
%\VignettePackage{LearnBioconductor}
%\VignetteEngine{knitr::knitr}
-->

```{r setup, echo=FALSE}
library(LearnBioconductor)
stopifnot(BiocInstaller::biocVersion() == "3.0")
```

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
knitr::opts_chunk$set(cache=TRUE, tidy=FALSE)
```

# Introduction to R

Martin Morgan<br/>
October 27, 2014

## R

Language and environment for statistical computing and graphics

- Full-featured programming language
- Interactive and *interpretted* -- convenient and forgiving
- Coherent, extensive documentation
- Statistical, e.g. `factor()`, `NA`
- Extensible -- CRAN, Bioconductor, github, ...

Vector, class, object

- Efficient _vectorized_ calculations on 'atomic' vectors `logical`,
  `integer`, `numeric`, `complex`, `character`, `byte`
- Atomic vectors are building blocks for more complicated _objects_
  - `matrix` -- atomic vector with 'dim' attribute
  - `data.frame` -- list of equal length atomic vectors
- Formal _classes_ represent complicated combinations of vectors,
  e.g., the return value of `lm()`, below

Function, generic, method

- Functions transform inputs to outputs, perhaps with side effects,
  e.g., `rnorm(1000)`
  - Argument matching first by name, then by position
  - Functions may define (some) arguments to have default values
- _Generic_ functions dispatch to specific _methods_ based on class of
  argument(s), e.g., `print()`. 
- Methods are functions that implement specific generics, e.g.,
  `print.factor`; methods are invoked _indirectly_, via the generic.

Introspection

- General properties, e.g., `class()`, `str()`
- Class-specific properties, e.g., `dim()`

Help

- `?print`: help on the generic print 
- `?print.data.frame`: help on print method for objects of class
    data.frame.

Example

```{r}
x <- rnorm(1000)                   # atomic vectors
y <- x + rnorm(1000, sd=.5)
df <- data.frame(x=x, y=y)         # object of class 'data.frame'
plot(y ~ x, df)                    # generic plot, method plot.formula
fit <- lm(y ~x, df)                # object of class 'lm'
methods(class=class(fit))          # introspection
```

## Lab

### 1. _R_ data manipulation

This exercise servers as a refresher / tutorial on basic input and
manipulation of data.

Input a file that contains ALL (acute lymphoblastic leukemia) patient
information

```{r echo=TRUE, eval=FALSE}
fname <- file.choose()   ## "ALLphenoData.tsv"
stopifnot(file.exists(fname))
pdata <- read.delim(fname)
```
```{r echo=FALSE}
fname <- system.file("extdata", "ALLphenoData.tsv", package="LearnBioconductor")
stopifnot(file.exists(fname))
pdata <- read.delim(fname)
```

Check out the help page `?read.delim` for input options, and explore
basic properties of the object you've created, for instance...

```{r ALL-properties}
class(pdata)
colnames(pdata)
dim(pdata)
head(pdata)
summary(pdata$sex)
summary(pdata$cyto.normal)
```

Remind yourselves about various ways to subset and access columns of a
data.frame

```{r ALL-subset}
pdata[1:5, 3:4]
pdata[1:5, ]
head(pdata[, 3:5])
tail(pdata[, 3:5], 3)
head(pdata$age)
head(pdata$sex)
head(pdata[pdata$age > 21,])
```

It seems from below that there are 17 females over 40 in the data set,
but when sub-setting `pdata` to contain just those individuals 19 rows
are selected. Why? What can we do to correct this?

```{r ALL-subset-NA}
idx <- pdata$sex == "F" & pdata$age > 40
table(idx)
dim(pdata[idx,])
```

Use the `mol.biol` column to subset the data to contain just
individuals with 'BCR/ABL' or 'NEG', e.g.,

```{r ALL-BCR/ABL-subset}
bcrabl <- pdata[pdata$mol.biol %in% c("BCR/ABL", "NEG"),]
```

The `mol.biol` column is a factor, and retains all levels even after
subsetting. How might you drop the unused factor levels?

```{r ALL-BCR/ABL-drop-unused}
bcrabl$mol.biol <- factor(bcrabl$mol.biol)
```

The `BT` column is a factor describing B- and T-cell subtypes

```{r ALL-BT}
levels(bcrabl$BT)
```

How might one collapse B1, B2, ... to a single type B, and likewise
for T1, T2, ..., so there are only two subtypes, B and T

```{r ALL-BT-recode}
table(bcrabl$BT)
levels(bcrabl$BT) <- substring(levels(bcrabl$BT), 1, 1)
table(bcrabl$BT)
```

Use `xtabs()` (cross-tabulation) to count the number of samples with
B- and T-cell types in each of the BCR/ABL and NEG groups

```{r ALL-BCR/ABL-BT}
xtabs(~ BT + mol.biol, bcrabl)
```

Use `aggregate()` to calculate the average age of males and females in
the BCR/ABL and NEG treatment groups.

```{r ALL-aggregate}
aggregate(age ~ mol.biol + sex, bcrabl, mean)
```

Use `t.test()` to compare the age of individuals in the BCR/ABL versus
NEG groups; visualize the results using `boxplot()`. In both cases,
use the `formula` interface. Consult the help page `?t.test` and re-do
the test assuming that variance of ages in the two groups is
identical. What parts of the test output change?

```{r ALL-age}
t.test(age ~ mol.biol, bcrabl)
boxplot(age ~ mol.biol, bcrabl)
```

## Resources

- [StackOverflow](http://stackoverflow.com/questions/tagged/r) for _R_
  programming questions; also [R-help]() mailing list.

Publications (General _R_)

<!-- Bibliography -->

[R]: http://r-project.org
