<!--
%\VignetteIndexEntry{01. Learning R / Bioconductor for Sequence Analysis}
%\VignettePackage{LearnBioconductor}
%\VignetteEngine{knitr::knitr}
-->

```{r setup, echo=FALSE}
library(LearnBioconductor)
stopifnot(BiocInstaller::biocVersion() == "3.0")
```

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# Learning R / Bioconductor for Sequence Analysis

This package contains training material for a Fall, 2014 introductory
R / Bioconductor course "Learning R / Bioconductor for Sequence
Analysis", offered October 27-29, Seattle, WA.

This course is directed at beginning and intermediate users who
would like an introduction to the analysis and comprehension of
high-throughput sequence data using R and Bioconductor. Day 1 focuses
on learning essential background: an introduction to the R programming
language; central concepts for effective use of Bioconductor software;
and an overview of high-throughput sequence analysis work flows. Day 2
emphasizes use of Bioconductor for specific tasks: an RNA-seq
differential expression work flow; exploratory, machine learning, and
other statistical tasks; gene set enrichment; and annotation.  Day 3
transitions to understanding effective approaches for managing larger
challenges: strategies for working with large data, writing re-usable
functions, developing reproducible reports and work flows, and
visualizing results.

The course combines lectures with extensive hands-on practicals;
students are required to bring a laptop with wireless internet access
and a modern version of the Chrome or Safari web browser.

## Schedule (tentative)

Day 1: Learn R / Bioconductor

- Introduction to R: objects, functions, help!
- Introduction to Bioconductor: working with packages and classes
- Introduction to sequence analysis: typical work flow; data types and
  quality assessment; essential Bioconductor packages

Day 2: Use R / Bioconductor

- An RNA-seq differential expression work flow (detail)
- Other work flows (survey): ChIP-seq, variants, copy number,
  epigenomics
- Machine learning; exploratory and other statistical analysis
- Annotating genes, genomes, and variants
- Approaches to gene set enrichment

Day 3: Develop Skills and Best Practices

- Working with large data
- Organzing code in functions, files, and packages
- Reproducible reports and work flows: markdown
- Visualization
