<!--
%\VignetteIndexEntry{03.2 Organizing Code in Functions, Files ,and Packages}
%\VignettePackage{LearnBioconductor}
%\VignetteEngine{knitr::knitr}
-->

```{r setup, echo=FALSE}
library(LearnBioconductor)
stopifnot(BiocInstaller::biocVersion() == "3.1")
```

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# Organizing Code in Functions, Files ,and Packages

Martin Morgan<br/>
February 3, 2015

## Programming

_R_ is a programming language

- Your own functions

  ```{r fun}
  fun <- function(n) {
      x <- rnorm(n)
      y <- x + rnorm(n, sd=.5)
      lm(y ~ x)
  }
  ```

    - Check out the _RStudio_ function wizard!

- Iteration

  ```{r iteration, eval=FALSE}
  ## 'side effects'
  for (i in 1:10)
      message(i)

  ## result for subsequent computation
  sapply(1:10, function(i) {
      sum(rnorm(1000))
  })
  ```
  
    - Other iteration paradigms: `lapply()`, `apply()`, `mapply() / Map()`.

- Conditional execution

  ```{r conditional, eval=FALSE}
  sapply(1:10, function(i) {
      if (sum(rnorm(1000)) > 0) {
          ans <- "hi!"
      } else {
          ans <- "low!"
      }
      ans
  })
  ```

## From script to package

Scripts contain a combination of data and transformations. Often the
transformations are idiosyncratic, and rely heavily on functions
provided by various packages. Sometimes the script contains a useful
chunk of code that could be reused in different places. Examples we've
encountered in this course might include GC-content of DNA sequences
(or is there a function for that already? check out
`r Biocpkg("Biostrings")`!) and creating a simple 'map' from one type of
annotation to another.

It is easy and beneficial to create a package. 

- Only need to think carefully about how to implement the function once.
- Code in the package is reused -- if you correct an error, then all
  your scripts automatically benefit.
- Share with your lab mates / colleagues for consistent results, and
  with the wider community for fame and glory.
  
What is an _R_ package?

- Simple directory structure of required files. Minimal:

        MyPackage
        |-- DESCRIPTION (title, author, version, etc.)
        |-- NAMESPACE (imports / exports)
        |-- R (R function definitions)
            |-- gc.R
            |-- annoHelper.R
        |-- man (help pages)
            |-- MyPackage.Rd
            |-- gc.Rd
            |-- annoHelper.Rd
  
Check out the _Rstudio_ package wizard!

## Lab

### GC-content

Write a function that takes a `DNAStringSet` and returns the GC content.

Modify the function using a conditional statement to work whether
provided a `DNAString` or a `DNAStringSet`. Test the function.

Save the function in a file on your AMI.

### Annotation-helper

Write a function that takes as its argument Ensembl gene identifiers
(like the `rownames()` of the _SummarizedExperiment_ object in the
RNASeq vignette yesterday) and uses the `select()` method and
annotation package `r Biocannopkg("org.Hs.eg.db")` to return a named
character vector, where the names of the vector are the Ensembl
identifiers and the values are the corresponding gene SYMBOLs. Adopt
some simple-to-implement policy for handling Ensembl identifiers that
map to more than one gene symbol. Save this function to another file

### A package

Use the _RStudio_ wizard to create a package from the files containing
your GC-content and annotation-helper functions.
