<!--
%\VignetteIndexEntry{03.4 Visualization}
%\VignettePackage{LearnBioconductor}
%\VignetteEngine{knitr::knitr}
-->

```{r setup, echo=FALSE}
library(LearnBioconductor)
stopifnot(BiocInstaller::biocVersion() == "3.1")
```

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# Visualization

Martin Morgan<br/>
February 3, 2015

[Three
approaches](http://bioconductor.org/help/course-materials/2014/summerx/Visualization-slides.pdf)

- 'base' graphics
- _lattice_
- `r CRANpkg("ggplot2")`
  
  - A neat _ggplot2_ example:
    [slide 19+](http://bioconductor.org/help/course-materials/2014/CSAMA2014/2_Tuesday/lectures/Visualization_in_Statistical_Genomics-Carey.pdf)
    from Vince Carey's presentation.


Genome-centric

- `r Biocpkg("Gviz")` for visualizing genomic regions
- `r Biocpkg("ggbio")` provides extensive options

Interactive visualization

- `r Biocpkg("rtracklayer")` for visualizing genomic ranges and
  managing UCSC genome browser sessions.
- `r Biocpkg('epivizr')` for interactive display
- `r CRANpkg("shiny")` [web](http://shiny.rstudio.com) for interactive
  apps.

Reports

- `r Biocpkg("ReportingTools")` for easy report templates

And...

- `r CRANpkg("RColorBrewer")` and
  [web site](http://colorbrewer2.org/): helps choose sensible color
  schemes.
- [Accidental aRt](http://accidental-art.tumblr.com/archive)
- [stick figure](https://github.com/EconometricsBySimulation/R-Graphics/blob/master/Stick-Figures/draw.stick.R)

## Lab

### Gviz / ggbio

Work through
[vignette section 2](http://bioconductor.org/packages/release/bioc/vignettes/Gviz/inst/doc/Gviz.pdf)
the `r Biocpkg('Gviz')` package, and then the 'Plotting fold changes
in genomic space' in the [RNASeq lab](B02.1.1_RNASeqLab.html).

Peruse the 
[vignette](http://bioconductor.org/packages/release/bioc/vignettes/ggbio/inst/doc/ggbio.pdf) of the `r Biocpkg('ggbio')` package.

Run the following _DESeq2_ work flow to arrive at a top-table; coerce
the result to a `data.frame`.

```{r ggplot-setup, eval=FALSE}
library(DESeq2)
library(airway)
data(airway)
se = airway
dds <- DESeqDataSet(se, design = ~ cell + dex)
dds$dex <- relevel(dds$dex, "untrt")
dds <- DESeq(dds)
res <- results(dds)
resdf <- as.data.frame(res)
```

A 'volcano plot' shows the relationship between P-value and log fold
change. Here's a basic volcano plot using base graphics; create a
volcano plot using _ggplot2_ and / or _lattice_.

```{r volcano, eval=FALSE}
plot(-log10(padj) ~ log2FoldChange, resdf)
```

